Automatic prostate cancer detection through \acs*{dce}-\acs*{mri} images: all you need is a good normalization

### 2016-08-24 02_methods/methods.tex
Finish to describe the results

\begin{figure}
  \centering
  \hspace*{\fill}
  \subfigure[Patient 1]{\label{subfig:pat1}\includegraphics[width=.49\textwidth]{02_methods/figures/pat1.pdf}} \hfill
  \subfigure[Patient 2]{\label{subfig:pat2}\includegraphics[width=.49\textwidth]{02_methods/figures/pat2.pdf}}
  \hspace*{\fill}
  \caption{Illustration of the variations of the intensity \acs*{pdf} over time of two patients in a \ac{dce}-\ac{mri}.}
  \label{fig:heatmap}
\end{figure}

In this work, we propose a method to normalize \ac{dce}-\ac{mri} prostate data, although it can be applied to any \ac{dce}-\ac{mri} sequences.
The aim of the method is to reduce the intra-patient variations, occurring during the acquisition process.
In \ac{t2w}-\ac{mri}, these variations imply that the intensity \ac{pdf} of the prostate region is shifted as well as wider or narrower, from a patient to another.
Therefore, these variations can be corrected using a $z$-score approach, assuming that the data follow a specific distribution~\citep{lemaitre2016normalization}.

In \ac{dce}-\ac{mri}, the inter-patient variations are more complex due to the over time acquisition.
These variations can be highlighted by observing the evolution of the intensity \ac{pdf} of the \ac{dce}-\ac{mri} over time, as shown in the heatmap in Fig.\,\ref{fig:heatmap}.
Therefore, these variations are:
(i) an intensity offset of the \ac{pdf} peak at pre-contrast,
(ii) a time offset depending of the contrast agent arrival, and
(iii) a change of scale related to the signal enhancement.
Therefore, our normalization method should attenuate all these variations and be performed globally across the different time sequence rather than for each independent sequence.

@@

Before to standardize each sequence, the first step of the normalization is to cancel the intensity specific at each patient, occurring due to the media injection.
In \ac{dce}-\ac{mri}, the intensity \ac{pdf} do not always follow either a Rician or a Gaussian distribution over time.
Therefore, the mean of these distributions cannot be used as a potential estimate for this offset.
Additionally, this offset should be characterized by a smooth transition between series over time.
Thus, this problem is solved using the graph-theory: considering the intensity \ac{pdf} over time as shown in Fig.\,\ref{fig:heatmap}, the offset is the boundary splitting the heatmap in two partitions such that it is as close as possible to the peak of the intensity \ac{pdf} (see Fig.\,\ref{fig:estimator} for an illustration).
Given the heatmap, a directed weighted graph $\mathcal{G}=(\mathcal{V}, \mathcal{E})$ is built by taking each bar of the heatmap as a node and connecting each pair of bars by an edge.
The edge weight $w_{ij}$ between two nodes $i$ and $j$ corresponding to two pixels at position $(x_i, y_i)$ and $(x_j, y_j)$, respectively, is defined as in Eq.\,\eqref{eq:weight}:

@@

@@

\noindent where $s(t)$ and $S_0$ are the \ac{mri} signal intensity at time $t$ and the average pre-contrast \ac{mri} signal intensity, respectively; $A$, $k_{el}$, and $k_{ep}$ are a constant proportional to the transfer constant, the diffusion rate constant, and the rate constant, respectively. Additionally, during the injection time $0 \leq t \leq \tau$, $t' = t$ and afterwards while $t > \tau$, $t' = \tau$.

Following this model, \citeauthor{hoffmann1995pharmacokinetic} propose the following similar model as expressed in Eq.\,\eqref{eq:hoffmann}:

\begin{equation}
  s_n(t) = 1 + \frac{A}{\tau} \left[ \frac{k_{ep} \left( \exp(k_{el} t') - 1 \right)}{k_{el}(k_{ep} - k_{el})} \exp(- k_{el} t) - \frac{\exp(k_{ep} t') - 1}{(k_{ep} - k_{el})} \exp(- k_{ep} t) \right].
  \label{eq:hoffmann}
\end{equation}

@@

\begin{description}
  \item[Segmentation of artery voxels and patient-based \ac{aif} estimation] The \ac{aif} signal from \ac{dce}-\ac{mri} can be manually estimated by selecting the most-enhanced voxels from the femoral or iliac arteries~\citep{meng2010comparison}.
    Few methods have been proposed to address the automated extraction of \ac{aif} signal.
    \citeauthor{chen2008automatic} filter successively the possible candidates~\citep{chen2008automatic}:
    (i) dynamic signals with small peak are rejecting by thresholding,
    (ii) voxels with a small wash-in are rejected by thresholding,
    (iii) a blob detector is used and large enough regions are kept, and
    (iv) circular and cylindricality are used to reject the last false positive.
    \citeauthor{zhu2011automated} propose an iterative method selecting voxels which best fit a gamma variate function~\citep{zhu2011automated}.
    However, it requires to compute first and second derivatives as well as maximum curvature points.
    \citeauthor{shanbhag2012generalized} propose a 4-steps algorithm~\citep{shanbhag2012generalized,fennessy2015quantitative}:
    (i) remove slices with artefacts and find the best slices based on intrinsic anatomic landmarks and enhancement characteristics,
    (ii) find the voxel candidates using the maximum enhanced voxels and a multi-label maximum entropy based thresholding algorithm,
    (iii) excluding region next to the endorectal coil, and
    (iv) selecting the best 5 candidates which meet enhancement characteristics and that are correlated.

@@

      $T_{10}$ can be estimated from the acquisition of a T$_1$ map.
      However, this modality was not part of the clinical trial in this research and the value of $T_{10}$ was fixed to \SI{1600}{\ms} for both blood and prostate as stated in the literature~\citep{fennessy2015quantitative,de2004mr,carr2011magnetic}.
      \item[Estimation of population-based \ac{aif}] While estimating the pharmacokinetic parameters from Tofts model, the \ac{aif} concentration $C_p(t)$ can be computed either from the patient or a population.
        We presented in the two previous sections the algorithms which allows to estimate the patient-based \ac{aif} concentration.
        To compare with the previous approach, we also computed a population-based \ac{aif} which will be also later used to compare the performance of both approaches.
        In that regard, the population-based \ac{aif} was estimated as in~\citep{meng2010comparison} by fitting the average patient-based \ac{aif}s to the model of~\cite{parker2006experimentally} which is formulated as in Eq.\,\eqref{eq:parker}:
        \begin{equation}
          C_p(t) = \sum_{n=1}^{2} \frac{A_n}{\sigma_n \sqrt{2 \pi}} \exp\left(\frac{- (t- T_n)^2}{2\sigma_{n}^{2}}\right) + \frac{\alpha \exp(-\beta t)}{1 + \exp{-s (t - \tau)}} ,
          \label{eq:parker}
        \end{equation}
        \noindent where $A_n$, $T_n$, and $\sigma_n$ are the scaling constants, centers, and widths of the n\textsuperscript{th} Gaussian, $\alpha$ and $\beta$ are the amplitude and decay constant of the exponential; and $s$ and $\tau$ are the width and center of the sigmoid function, respectively.
\end{description}

>>>>

\begin{figure}
  \centering
  \includegraphics[width=0.7\linewidth]{02_methods/figures/t2w.pdf}
  \caption{Illustration of the inter-patient variations in 17 different patients, using the \acs*{pdf} representation.}
  \label{fig:t2w}
\end{figure}

In this work, we propose a method to normalize \ac{dce}-\ac{mri} prostate data to reduce intra-patient variations, although it can be applied to any \ac{dce}-\ac{mri} sequences.
In \ac{t2w}-\ac{mri}, these variations are characterized by a shift and a scaling of the intensities as illustrated by the intensity \ac{pdf} in Fig.\,\ref{fig:t2w}.
Therefore, these variations can be corrected using a $z$-score approach, assuming that the data follow a specific distribution~\citep{lemaitre2016normalization}.

\begin{figure}
  \centering
  \hspace*{\fill}
  \subfigure[]{\label{subfig:pathist}\includegraphics[width=.35\textwidth]{02_methods/figures/IMG_20160824_122328.jpg}} \hfill
  \subfigure[]{\label{subfig:heatmaprep}\includegraphics[width=.5\textwidth]{02_methods/figures/pat2.pdf}}
  \hspace*{\fill}
  \\
  \hspace*{\fill}
  \subfigure{\includegraphics[width=.2\textwidth]{02_methods/figures/pat1.pdf}} \hfill
  \subfigure[]{\label{subfig:pat2}\includegraphics[width=.2\textwidth]{02_methods/figures/pat2.pdf}} \hfill
  \subfigure{\includegraphics[width=.2\textwidth]{02_methods/figures/pat2.pdf}}
  \hspace*{\fill}
  \caption{Illustration of the variations of the intensity \acs*{pdf} over time of two patients in a \ac{dce}-\ac{mri}.}
  \label{fig:heatmap}
\end{figure}

In \ac{dce}-\ac{mri}, the intensity \ac{pdf} does not follow a unique type of distribution such as Rician or Gaussian distribution, as shown in Fig.\,\ref{subfig:pathist}.
Indeed, the inter-patient variations are more complex due to the temporal acquisition.
A better representation to observed these variations is to represent the intensity \ac{pdf} of the \ac{dce}-\ac{mri} data over time (see Fig.\,\ref{subfig:pathist}) using a heatmap representation as shown in Fig.\,\ref{subfig:heatmaprep}.
Analyzing this heatmap representation across patients (see Fig.\,\ref{subfig:pat2}), the following variations are highlighted:
(i) an intensity offset $\Delta_i$ of the \ac{pdf} peak at pre-contrast,
(ii) a time offset $\Delta_t$ depending of the contrast agent arrival, and
(iii) a change of scale $\sigma_i$ related to the signal enhancement.
Therefore, our normalization method should attenuate all these variations and be performed globally across the different time sequence rather than for each independent sequence.

@@

Before to standardize each sequence, the first step of the normalization is to cancel the intensity specific at each patient, occurring due to the media injection.
As previously mentioned, the intensity \ac{pdf} do not always follow either a Rician or a Gaussian distribution over time, in \ac{dce}-\ac{mri}.
Therefore, the mean of these distributions cannot be used as a potential estimate for this offset.
Additionally, this offset should be characterized by a smooth transition between series over time.
Thus, this problem is solved using the graph-theory: considering the intensity \ac{pdf} over time as shown in Fig.\,\ref{subfig:heatmaprep}, the offset is the boundary splitting the heatmap in two partitions such that it is as close as possible to the peak of the intensity \ac{pdf} (see Fig.\,\ref{fig:estimator} for an illustration).
Given the heatmap, a directed weighted graph $\mathcal{G}=(\mathcal{V}, \mathcal{E})$ is built by taking each bar of the heatmap as a node and connecting each pair of bars by an edge.
The edge weight $w_{ij}$ between two nodes $i$ and $j$ corresponding to two pixels at position $(x_i, y_i)$ and $(x_j, y_j)$, respectively, is defined as in Eq.\,\eqref{eq:weight}:

@@

As stated in the introduction, one of our main challenge is to demonstrate the benefit of using our normalization method prior to extract parameters using quantitative and semi-quantitative methods.
In this section, we summarize the different methods which have been used for the quantification of \ac{dce}-\ac{mri} for \ac{cap} detection~\citep{lemaitre2015computer}.
Furthermore, we would like to emphasize the following additional contributions: (i) a novel automatic \ac{aif} estimation algorithm based on clustering and (ii) a simplified semi-quantitative method using constrained optimization.

@@

\noindent where $s(t)$ and $S_0$ are the \ac{mri} signal intensity at time $t$ and the average pre-contrast \ac{mri} signal intensity, respectively; $A$, $k_{el}$, and $k_{ep}$ are a constant proportional to the transfer constant, the diffusion rate constant, and the rate constant, respectively.
Additionally, $t'$ is set such that $0 \leq t \leq \tau$, $t' = t$ and afterwards while $t > \tau$, $t' = \tau$.

\citeauthor{hoffmann1995pharmacokinetic} propose a similar model as expressed in Eq.\,\eqref{eq:hoffmann}, which derive from the Brix model:

\begin{equation}
  \small
  s_n(t) = 1 + \frac{A}{\tau} \left[ \frac{k_{ep} \left( \exp(k_{el} t') - 1 \right)}{k_{el}(k_{ep} - k_{el})} \exp(- k_{el} t) - \frac{\exp(k_{ep} t') - 1}{(k_{ep} - k_{el})} \exp(- k_{ep} t) \right].
  \label{eq:hoffmann}
\end{equation}

The parameters $A$, $k_{el}$, and $k_{ep}$ are estimated by fitting the model using non-linear least-squares optimization solved with Levenberg-Marcquardt.

@@

\begin{description}
  \item[Segmentation of artery voxels and patient-based \ac{aif} estimation] The \ac{aif} signal from \ac{dce}-\ac{mri} can be manually estimated by selecting the most-enhanced voxels from the femoral or iliac arteries~\citep{meng2010comparison}.
    Few methods have been proposed to address the automated extraction of \ac{aif} signal.
    \citeauthor{chen2008automatic} filter successively the possible candidates~\citep{chen2008automatic}:
    (i) dynamic signals with small peak are rejecting by thresholding,
    (ii) voxels with a small wash-in are rejected by thresholding,
    (iii) a blob detector is used and large enough regions are kept, and
    (iv) circular and cylindricality criteria are used to reject the last false positive.
    \citeauthor{zhu2011automated} propose an iterative method selecting voxels which best fit a gamma variate function~\citep{zhu2011automated}.
    However, it requires to compute first and second derivatives as well as maximum curvature points.
    \citeauthor{shanbhag2012generalized} propose a 4-steps algorithm~\citep{shanbhag2012generalized,fennessy2015quantitative}:
    (i) remove slices with artefacts and find the best slices based on intrinsic anatomic landmarks and enhancement characteristics,
    (ii) find the voxel candidates using the maximum enhanced voxels and a multi-label maximum entropy based thresholding algorithm,
    (iii) excluding region next to the endorectal coil, and
    (iv) selecting the best 5 candidates which meet enhancement characteristics and that are correlated.

@@

      $T_{10}$ can be estimated from the acquisition of a T$_1$ map.
      However, this modality is not part of the clinical trial in this research and the value of $T_{10}$ is fixed to \SI{1600}{\ms} for both blood and prostate, in accordance with the values found in the literature~\citep{fennessy2015quantitative,de2004mr,carr2011magnetic}.
      \item[Estimation of population-based \ac{aif}] While estimating the pharmacokinetic parameters from Tofts model, the \ac{aif} concentration $C_p(t)$ can be computed either from the patient or a population.
        We presented in the two previous sections the algorithms which allows to estimate the patient-based \ac{aif} concentration.
        To compare with the previous approach, we also computed a population-based \ac{aif} which will be also later used to compare the performance of both approaches.
        In that regard, the population-based \ac{aif} was estimated as in~\citep{meng2010comparison} by fitting the average patient-based \ac{aif}s to the model of~\cite{parker2006experimentally} which is formulated as in Eq.\,\eqref{eq:parker}:
        \begin{equation}
          C_p(t) = \sum_{n=1}^{2} \frac{A_n}{\sigma_n \sqrt{2 \pi}} \exp\left(\frac{- (t- T_n)^2}{2\sigma_{n}^{2}}\right) + \frac{\alpha \exp(-\beta t)}{1 + \exp{-s (t - \tau)}} ,
          \label{eq:parker}
        \end{equation}
        \noindent where $A_n$, $T_n$, and $\sigma_n$ are the scaling constants, centers, and widths of the n\textsuperscript{th} Gaussian, $\alpha$ and $\beta$ are the amplitude and decay constant of the exponential; and $s$ and $\tau$ are the width and center of the sigmoid function, respectively.
\end{description}

### 2016-08-24 03_experiments/experiments.tex
Finish to describe the results

\ac{dce}-\ac{mri} are performed using a fat suppressed 3D T$_1$ VIBE sequence (\ac{tr}/\ac{te}/Flip angle: \SI{3.25}{\ms}/\SI{1.12}{\ms}/\SI{10}{\degree}; Matrix: $256 \times 192$; \ac{fov}: $280 \times 210$ (with 75\% rectangular \ac{fov}); slab of 16 partitions of \SI{3.5}{\mm} thickness; temporal resolution: \SI{6}{\s}/slab over approximately \SI{5}{\minute}).
A power injector (Medrad, Indianola, USA) is used to provide a bolus injection of Gd-DTPA (Dotarem, Guerbet, Roissy, France) at a dose of \SI{0.2}{\ml} Gd-DTPA/kg of body weight.

These \ac{dce}-\ac{mri} sequences are resampled using the spatial information of the \ac{t2w}-\ac{mri} and missing data are interpolated using linear interpolation.
The volumes of the \ac{dce}-\ac{mri} dynamic are rigidly registered, to remove any patient motion during the acquisition.
Furthermore, a non-rigid registration is performed between the \ac{t2w}-\ac{mri} and \ac{dce}-\ac{mri} in order to propagate the prostate zones and \ac{cap} ground-truths.
The resampling is implemented in C++ using the Insight Segmentation and Registration Toolkit~\citep{ibanez2005itk}.

@@

The first experiment consists in comparing the classification performance using each individual parameter for the different models, to assess the potential benefit of the normalization method.
Therefore, each feature is individually classified using a Gaussian \ac{nb} classifier in a \ac{lopo} fashion.
{\color{red}\ac{nb} is used due to its simplicity and it enables to check the fitted parameters for interpretation.}
The results are summarized in Table~\ref{tab:resfeats}.
It can be noted that in the majority of the cases, normalizing the data improve the classification performance in terms of \ac{auc}.
Only the \ac{pun} model does not follow this tendency {\color{red} which might due to the fact that the function do not fit the data as good as the other model - WE WILL NEED SOME RMSE FOR EACH FITTING IF WE SAY SO.}.

{\color{red}It could be interesting to check the value of the mean and std of the \ac{nb}. We could compute a ratio which refer to the class separability.}

@@

@@

Usually, the parameters are combined to form a multi-dimensional prior to classify the data.
Therefore, for each model, all parameters are combined and classified using a \ac{rf} classifier in a \ac{lopo} fashion.
The use of \ac{rf} is motivated since that it leads to the best performance in the state-of-the-art methods~\citep{litjens2014computer}.
The results are summarized by making \ac{roc} analysis and computing the \ac{auc}, as reported in Fig.\,\ref{fig:normpharmarf}.
It can be noted that in all the cases the normalization of the data improve the classification performance.

@@

@@

The quantification methods are extracting a minimal set of parameters which should characterized the enhancement \ac{dce}-\ac{mri} curves.
However, this extraction could discard some information of the signal.
This experiment attends to use the whole \ac{dce}-\ac{mri} signal to perform the classification.
Therefore, the enhanced signal is classified using a \ac{rf} classifier in a \ac{lopo} fashion.
The \ac{roc} analysis and \ac{auc} are reported in Fig.\,\ref{fig:rfnormdcesignal}.
It can be note that the worst performance are achieved while data without normalization.
However, normalizing the data, this classification strategy lead the best classification performance, outperforming any quantification method, showing the importance of our normalization algorithm.

@@

>>>>

\ac{dce}-\ac{mri} is performed using a fat suppressed 3D T$_1$ VIBE sequence (\ac{tr}/\ac{te}/Flip angle: \SI{3.25}{\ms}/\SI{1.12}{\ms}/\SI{10}{\degree}; Matrix: $256 \times 192$; \ac{fov}: $280 \times 210$ (with 75\% rectangular \ac{fov}); slab of 16 partitions of \SI{3.5}{\mm} thickness; temporal resolution: \SI{6}{\s}/slab over approximately \SI{5}{\minute}).
A power injector (Medrad, Indianola, USA) is used to provide a bolus injection of Gd-DTPA (Dotarem, Guerbet, Roissy, France) at a dose of \SI{0.2}{\ml} Gd-DTPA/kg of body weight.

These \ac{dce}-\ac{mri} sequences are resampled using the spatial information of the \ac{t2w}-\ac{mri} and missing data are interpolated using a linear interpolation.
The volumes of the \ac{dce}-\ac{mri} dynamic are rigidly registered, to remove any patient motion during the acquisition.
Furthermore, a non-rigid registration is performed between the \ac{t2w}-\ac{mri} and \ac{dce}-\ac{mri} in order to propagate the prostate zones and \ac{cap} ground-truths.
The resampling is implemented in C++ using the Insight Segmentation and Registration Toolkit~\citep{ibanez2005itk}.

@@

@@

This experiment consists in assessing the benefit of normalizing the \ac{dce}-\ac{mri}, prior to classify each individual parameter for each model.
Therefore, a Gaussian \ac{nb} classifier is used in conjunction with a \ac{lopo}.
The results are summarized in Table~\ref{tab:resfeats}.
In general, data normalization leads to better classification performance while extracting pharmacokinetic related parameters.
Only the \ac{pun} model does not follow this tendency.
{\color{red} Find out why!!!}

@@

@@

In previous proposed \ac{cad} systems, the pharmacokinetic of a single model are combined.
Therefore, for each model, all parameters are combined and classified using a \ac{rf} classifier in a \ac{lopo} fashion.
The use of \ac{rf} is motivated since that it leads to the best performance in the state-of-the-art methods~\citep{litjens2014computer}.
Results are summarized by performing a \ac{roc} analysis and computing the \ac{auc}, as reported in Fig.\,\ref{fig:normpharmarf}.
Independently to the quantification model, the data normalization improves the\ac{auc} and classification performance.

@@

@@

As stated in the introduction, the quantification methods are extracting a set of parameters characterizing the enhancement \ac{dce}-\ac{mri} signal.
However, this extraction could lead to a loss of information.
This experiment is performed to assess if making use of the whole \ac{dce}-\ac{mri} signal instead of the just the pharmacokinetic parameters can improve the classification performance.
Therefore, each enhanced \ac{dce}-\ac{mri} signal, normalized and un-normalized, is classified using a \ac{rf} classifier in a \ac{lopo} fashion.
The \ac{roc} analysis and \ac{auc} are reported in Fig.\,\ref{fig:rfnormdcesignal}.
It can be noted that the worst performance are achieved while classifying raw data without normalization.
However, data normalization in conjunction with the use of the whole \ac{dce}-\ac{mri} signal is the strategy which outperforms all others.
